Automatic discovery of cell types and microcircuitry from neural connectomics

### 2014-06-23 docs/paper.tex
incorporated yarden's first set of suggestions

\begin{abstract}
  Neural connectomics has begun producing massive amounts of data,
  necessitating new analysis methods to discover the biological and
  computational structure. It has long been assumed that discovering
  neuron types and their relation to microcircuitry is crucial to
  understanding neural function. Here we developed a nonparametric
  Bayesian technique that identifies neuron types and microcircuitry
  patterns in connectomics data. It combines the information
  traditionally used by biologists, including connectivity, cell body
  location and the spatial distribution of synapses.  We show that the
  approach recovers known neuron types in the retina and enables
  predictions of connectivity, better than simpler algorithms. It also
  can reveal interesting structure in the nervous system of
  c. elegans, and automatically discovers the structure of a
  microprocessor.  Our approach extracts structural meaning from
  connectomics, enabling new approaches of automatically deriving
  anatomical insights from these emerging datasets.
\end{abstract}

\section*{Introduction}
Emerging connectomics techniques \autocite{Morgan2013,Zador2012}
promise to quantify the location and connectivity of each neuron
within a tissue volume. These massive datasets will far exceed the
capacity of neuroanatomists to manually trace small circuits, thus
necessitating computational, quantitative, and automatic methods for
understanding neural structure.  The impact of this kind of
high-throughput transition has been seen before -- rise of sequencing
techniques necessitated the development of novel computational methods
to understand genomic structure, ushering in an era of bioinformatics
as an independent discipline \autocite{Koboldt2013}.

The brain consists of multiple kinds of neurons, each of which is
hypothesized to have a specific role in the overall
computation. Neuron types differ in many ways, e.g. chemical or
morphological, but they also differ in the way they connect to one
another. In fact, the idea of well defined, type-dependent local
connectivity patterns (microcircuits) has been prominent in many
areas, ranging from sensory (e.g. retina, \autocite{Masland2001} via
processing (e.g. neocortex \autocite{Mountcastle1997}) to movement
(e.g. spinal cord) \autocite{Grillner2005}. These sorts of repeated
computing patterns are a common feature of computing systems, even
arising in human-made computing circuits. It remains an important
challenge to develop algorithms to use anatomical data,
e.g. connectomics, to automatically back out underlying
microcircuitry.

@@

The discovery of structure in probabilistic graphs is a crucial aspect
in machine learning. Commonly used algorithms include
community-based-detection methods \autocite{Girvan2002}, and
stochastic block models \autocite{Nowicki2001}.  While these
approaches can incorporate the probabilistic nature of neural
connections \autocite{Hill2012} they do not incorporate the additional
richer structure present in connectomics data -- the location of cell
bodies, the spatial distribution of synapses, and the distances
between neurons. Of particular importance is that the probability of
connections has a strong spatial component, a factor that is hard to
reconcile with many other methods. A model attempting to fully capture
the variation in the nervous system should take into account the broad
set of available features.

@@

We primarily focus on the recently-released mouse retina connectome
\autocite{Helmstaedter2013}, but additionally examine the c. elegans
connectome \autocite{White1986}, and then ``connectome'' of a
classical microprocessor \autocite{James2010}. Comparing the cell
types discovered by the algorithms with those obtained manually by
human anatomists reveals a high degree of agreement. We present a
scalable approach to infer microcircuitry from connectomics data.

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=183mm]{fig1b.ai}}
  \caption{An overview of our method. a. Input data is a connectivity
    matrix indicating whether cell $i$ (row) synapses on cell $j$
    (column), a distance matrix between cell soma, b. the depth of the
    cell soma, and the c. depth profile of the cells synaptic
    connectivity. d. Our algorithm discovers hidden cell types in this
    connectivity data by assuming all cells of a type share a
    distance-dependent connectivity profile, similar depth, and a
    similar synpatic density profile, with cells of other types.  This
    results in a clustering of the cells by those hidden
    types. d. shows the cell connectivity matrix with cells of the
    same type grouped together. e.) shows the learned probability of
    connection between our different types at various distances -- in
    this case, the cells are likely to connect when they are
    close. f. shows a connectivity profile with virtually no
    probability of connectivity. We can plot this these types g.) to
    show the relationship and spatial connections, and how probable
    various types are to connect to one another. h. The connectivity
    between our discovered types changes as a function of distance. }

@@

We start with the generic assumption that cell types exist -- a cell
is a member of a single, unobserved type -- and these types affect
soma depth, distribution of synapses, as well as a cell type and
distance dependent connection probability. From this assumption we
develop a Bayesian algorithm that estimates the underlying cell types.
We start with a model for connectivity, the infinite stochastic block
model (iSBM)\autocite{Kemp2006a,Xu2006}, which has been shown to
meaningfully cluster connection graphs while learning the number of
hidden groups, or types. We extend this approach by adding distance
dependence to model salient aspects of microcircuitry. We additionally
model cell body depth as a mixture of Gaussians and each cell’s
synapse density profile as mixture of mixtures of Gaussians.

For example, retinal ganglion cells may synapse on nearby, but not far
away, amacrine cells, with ganglion cells being superficial and having
a broad distribution of synapses.  This procedure automatically learns
the number of cell types, which cells belong to which types, along
with the distance and type dependent connectivity probabilities (see
fig~\ref{fig:overview}). We apply our algorithm to datasets from mouse
retina, c-elegans and a historical microprocessor. Anatomists classify
cells based on many features, giving us a meaningful baseline to
compare against.

Before we can use a probabilistic model to analyze real data we first
need to assess if the model works properly. We thus simulate data for
which we know the correct structure and comparing the estimated
structure based on the algorithm (see methods) with the one we used
for simulation. We find that the model does a good job of recovering
the correct number of cell types, (fig~\ref{fig:synthetic}a), the the
cell identities (fig~\ref{fig:synthetic}b), and the spatial extent of
each type (fig~\ref{fig:synthetic}c).  Indeed, it does far better than
the existing infinite stochastic block model that assumes that there
is no distance dependence to the connectivity. The model converges
relatively quickly to an estimate of the most probable values for the
cell types, which is enabled by using a combination of simulated
annealing and parallelized Markov chain Monte Carlo (see methods for
details). It appears that the model behaves the way it should and
allows application to relatively large datasets.

@@

The roundworm Caenorhabditis elegans is a model system in
developmental neuroscience\autocite{White1986}, with the location and
connectivity of each of 302 neurons developmentally determined,
leading to early measurement of the connectome. Unlike the retina,
only the motor neurons in c. elegans exhibit regular distribution in
space -- along the body axis. Most interneurons are
concentrated in various ganglia that project throughout the entire
animal, and the sensory neurons are primarily located in a small
number of anterior ganglia.

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=7in]{otherdata.ai}}
  \caption{Discovering connectivity and type in both
    \textit{c. elegans} and the MOS 6502 microprocessor.  a.) Initial
    \textit{c. elegans} cell connectivity matrix, red are
    (directional) chemical synapses, blue are electrical
    synapses. Point size is proportional to synapse count between
    those cells. b.) Clustered adjacency matrix showing identified
    clusters. c.) Inter-cluster connectivity, showing the position of
    each cell along the body axis as well as the ``true'' cell type --
    both coarse (motor, sensory, and interneuron) as well as fine
    (small text labels). d.) The clusters and connectivity versus
    distance for connections between Gate and C1, Gate and C2, and C1
    and C2 terminals on a transistor. Red and teal types have a
    terminal pulled down to ground and mostly function as
    inverters. Purple class are clocked, stateful transistors, orange
    control the ALU and yellow control the special data bus. e.) The
    spatial distribution of the transistors in each cluster.}
  \label{fig:otherdata}
\end{figure}

@@

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=60mm]{mixing.ari.pdf}}
  \caption{}
\label{fig:mixing:ari}
\end{figure}

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=130mm]{mixing.scorevstime.pdf}}
  \caption{}
\label{fig:mixing:scorevstime}
\end{figure}

>>>>

\begin{abstract}
  Neural connectomics has begun producing massive amounts of data,
  necessitating new analysis methods to discover the biological and
  computational structure. It has long been assumed that discovering
  neuron types and their relation to microcircuitry is crucial to
  understanding neural function. Here we developed a nonparametric
  Bayesian technique that identifies neuron types and microcircuitry
  patterns in connectomics data. It combines the information
  traditionally used by biologists, including connectivity, cell body
  location and the spatial distribution of synapses, in a
  principled and probabilstically-coherent manner. We show that the
  approach recovers known neuron types in the retina and enables
  predictions of connectivity, better than simpler algorithms. It also
  can reveal interesting structure in the nervous system of
  C. elegans, and automatically discovers the structure of a
  microprocessor.  Our approach extracts structural meaning from
  connectomics, enabling new approaches of automatically deriving
  anatomical insights from these emerging datasets.
\end{abstract}

\section*{Introduction}
Emerging connectomics techniques \autocite{Morgan2013,Zador2012}
promise to quantify the location and connectivity of each neuron
within a tissue volume. These massive datasets will far exceed the
capacity of neuroanatomists to manually trace small circuits, thus
necessitating computational, quantitative, and automatic methods for
understanding neural circuit structure.  The impact of this kind of
high-throughput transition has been seen before -- rise of sequencing
techniques necessitated the development of novel computational methods
to understand genomic structure, ushering in  bioinformatics
as an independent discipline \autocite{Koboldt2013}.

The brain consists of multiple kinds of neurons, each of which is
hypothesized to have a specific role in the overall
computation. Neuron types differ in many ways, e.g. chemical or
morphological, but they also differ in the way they connect to one
another. In fact, the idea of well defined, type-dependent local
connectivity patterns (microcircuits) has been prominent in many
areas, from sensory (e.g. retina, \autocite{Masland2001} to processing
(e.g. neocortex \autocite{Mountcastle1997}) to movement (e.g. spinal
cord) \autocite{Grillner2005}. These sorts of repeated computing
patterns are a common feature of computing systems, even arising in
human-made computing circuits. It remains an important challenge to
develop algorithms to use anatomical data, e.g. connectomics, to
automatically back out underlying microcircuitry.

@@

The discovery of structure in probabilistic graphs is a well-known
problem in machine learning. Commonly used algorithms include
community-based-detection methods \autocite{Girvan2002}, and
stochastic block models \autocite{Nowicki2001}.  While these
approaches can incorporate the probabilistic nature of neural
connections \autocite{Hill2012} they do not incorporate the additional
richer structure present in connectomics data -- the location of cell
bodies, the spatial distribution of synapses, and the distances
between neurons. Of particular importance is that the probability of
connections has a strong spatial component, a factor that is hard to
reconcile with many other methods. A model attempting to fully capture
the variation in the nervous system should take into account the broad
set of available features.

@@

We primarily focus on the recently-released mouse retina connectome
\autocite{Helmstaedter2013}, but additionally examine the C. elegans
connectome \autocite{White1986}, and then ``connectome'' of a
classical microprocessor \autocite{James2010}. Comparing the cell
types discovered by the algorithms with those obtained manually by
human anatomists reveals a high degree of agreement. We thus present a
scalable probabilistic approach to infer microcircuitry from
connectomics data available today and in the future.

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=183mm]{fig1b.ai}}
  \caption{An overview of our method. a. Input data is a connectivity
    matrix indicating whether cell $i$ (row) synapses on cell $j$
    (column), a distance matrix between cell soma, b. the depth of the
    cell soma, and the c. depth profile of the cells synaptic
    connectivity. d. Our algorithm discovers hidden cell types in this
    connectivity data by assuming all cells of a type share a
    distance-dependent connectivity profile, similar depth, and a
    similar synpatic density profile, with cells of other types.  This
    results in a clustering of the cells by those hidden
    types. d. shows the cell connectivity matrix with cells of the
    same type grouped together. e. shows the learned probability of
    connection between our different types at various distances -- in
    this case, the cells are likely to connect when they are
    close. f. shows the probability of connection between two cell types
that very rarely connect -- there's a background ``base'' connection rate
to account for errors in data, but the probability is very low. 
We can plot this these types g.) to show the relationship and spatial
connections, and how probable various types are to connect to one
another. h. The connectivity between our discovered types changes as a
function of distance. }

@@

We build a structured probabilistic model which begins with the
generic notion of a cell being a member of a single, unobserved type
-- and these types affect soma depth, distribution of synapses, as
well as a cell type and distance dependent connection probability. For
example, retinal ganglion cells may synapse on nearby, but not far
away, amacrine cells, with ganglion cells being superficial and having
a broad distribution of synapses.

From these assumptions (priors) develop a generative Bayesian model
that estimates the underlying cell types and how they connect. We take
as input (fig ~\ref{fig:overview}a) the connectivity matrix of cells,
a matrix of the distance between cells, and the per-cell soma depth
(fig ~\ref{fig:overview}b) and the depth profile of the cell's
synapses (fig ~\ref{fig:overview}c). We perform joint probabilistic
inference to automatically learn the number of cell types, which cells
belong to which type, their type-specific connectivity, and how
connections between types vary with distance. We also simultaneously
learn the soma depth associated with each type and the typical
synaptic density profile (fig ~\ref{fig:overview}d.).

We apply our algorithm to datasets from mouse
retina, C. elegans and a historical microprocessor. Anatomists
classify cells based on many features, giving us a meaningful baseline
to compare against.

We start with a model for connectivity, the infinite stochastic block
model (iSBM)\autocite{Kemp2006a,Xu2006}, which has been shown to
meaningfully cluster connection graphs while learning the number of
hidden groups, or types. We extend this approach by adding distance
dependence to model salient aspects of microcircuitry via logistic and
exponential distance-link functions.  We additionally model cell body
depth unimodally and synapse density profile multimodially (see
Methods for mathematical details).

To validate our model, we performed a series of simulations to test if
the model can accurately recover the true underlying network structure
and cell type identity.  We thus simulate data for which we know the
correct structure and comparing the estimated structure based on the
algorithm (see methods) with the one we used for simulation. We find
that the model does a good job of recovering the correct number of
cell types, (fig~\ref{fig:synthetic}a), the the cell identities
(fig~\ref{fig:synthetic}b), and the spatial extent of each type
(fig~\ref{fig:synthetic}c).  Indeed, it does far better than the
existing infinite stochastic block model that assumes that there is no
distance dependence to the connectivity. The model converges
relatively quickly to an estimate of the most probable values for the
cell types, which is enabled by using a combination of simulated
annealing and parallelized Markov chain Monte Carlo (see methods for
details). Thus we can apply our model to simulated datasets with structure
and scale similar to that of our biological datasets and recover the known 
correct structure.

@@

The roundworm Caenorhabditis elegans is a model system in
developmental neuroscience\autocite{White1986}, with the location and
connectivity of each of 302 neurons developmentally determined,
leading to early measurement of the connectome. Unlike the retina,
only the motor neurons in C. elegans exhibit regular distribution in
space -- along the body axis. Most interneurons are
concentrated in various ganglia that project throughout the entire
animal, and the sensory neurons are primarily located in a small
number of anterior ganglia.

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=7in]{otherdata.ai}}
  \caption{Discovering connectivity and type in both
    \textit{C. elegans} and the MOS 6502 microprocessor.  a.) Initial
    \textit{C. elegans} cell connectivity matrix, red are
    (directional) chemical synapses, blue are electrical
    synapses. Point size is proportional to synapse count between
    those cells. b.) Clustered adjacency matrix showing identified
    clusters. c.) Inter-cluster connectivity, showing the position of
    each cell along the body axis as well as the ``true'' cell type --
    both coarse (motor, sensory, and interneuron) as well as fine
    (small text labels). d.) The clusters and connectivity versus
    distance for connections between Gate and C1, Gate and C2, and C1
    and C2 terminals on a transistor. Red and teal types have a
    terminal pulled down to ground and mostly function as
    inverters. Purple class are clocked, stateful transistors, orange
    control the ALU and yellow control the special data bus. e.) The
    spatial distribution of the transistors in each cluster.}
  \label{fig:otherdata}
\end{figure}

@@

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=60mm]{mixing.ari.pdf}}
  \caption{Adjusted rand index for synthetic data as a function of run iteration. }
\label{fig:mixing:ari}
\end{figure}

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=130mm]{mixing.scorevstime.pdf}}
  \caption{Total model score (log score) vs iteration}
\label{fig:mixing:scorevstime}
\end{figure}
